---
title: Pre-workshop technical Setup🖥️
format: html
date: 2024/04/10
date-format: long
date-modified: last-modified
editor: 
  markdown: 
    wrap: sentence
---

```{r}
#| label: chunk-setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Pre-workshop technical instructions for participants.
Let's get ready to rock!
🚀  

## Required Software Checklist ✅

Before attending SSoQE 2025, please ensure you have the following software installed:

- [ ] **R** (version 4.5+ recommended)
- [ ] **RStudio** (latest version)
- [ ] **Rtools** (Windows only)
- [ ] **Git** (for version control)
- [ ] **GitHub account** (free)

## R and RStudio 💻

- [R](https://en.wikipedia.org/wiki/R_(programming_language)) is a fantastic software for statistical analyses.
- 📊 [RStudio](https://posit.co/products/open-source/rstudio/) is your trusty sidekick, helping you navigate the R universe with ease.

### Install/Update R

Make sure you have **R version 4.5 or later** installed. You can:

- Download the latest version from [CRAN](https://cran.r-project.org/)
- Check your current version by running `R.version.string` in R Console

```r
R.version.string
# [1] "R version 4.5.1 (2025-06-13 ucrt)"
```

### Install/Update RStudio

Download and install the **latest version** of RStudio from the [RStudio website](https://rstudio.com/products/rstudio/download/).

You can check your RStudio version by running:

```r
rstudioapi::versionInfo()$version
# [1] ‘2025.5.1.513’
```

### Rtools (Windows Users Only) 🪟

If you're using Windows, you'll need **Rtools** for building R packages. Download the version compatible with your R version from the [CRAN Rtools page](https://cran.r-project.org/bin/windows/Rtools/).

For R 4.5+, use Rtools 4.5 etc.

### Additional Resources

There are plenty of additional guides available to help you:

-   [R Basics for Paleoecologists](https://ckiahtipes.github.io/) by C.A. Kiahtipes
-   [Install or Update R tutorial](https://jennhuck.github.io/workshops/install_update_R.html) by Jennifer Huck



## Git and GitHub 🌐

Please follow the tutorial [here](https://ondrejmottl.github.io/VersionControl_gitHell/) to install Git and set up a GitHub account.

Specifically, we will be using Git, GitHub, and RStudio for version control and collaboration during the workshop.
🤝

### Quick setup checklist

1. **Install Git** - Download from [git-scm.com](https://git-scm.com/)
2. **Create GitHub account** - Sign up at [github.com](https://github.com/)
3. **Configure Git** with your name and email:
   ```bash
   git config --global user.name "Your Name"
   git config --global user.email "your.email@example.com"
   ```
4. **Set up authentication** - We'll help you with Personal Access Tokens (PAT) during the workshop

## Packages 📦

Packages are like magic toolboxes 🧰 that contain a collection of functions for specific needs.
We want to make sure that everyone has the necessary packages installed for this workshop.

### Essential packages for SSoQE 2025

Let's create a list of the core packages we'll need from CRAN:

```{r}
#| label: chunk-pkg-list
#| output: false
#| warning: false
#| message: false
# Complete list of R packages used across SSoQE lectures
# Generated on September 11, 2025
ssoqe_package_list <-
  c(
    # Project management & Environment
    "here",
    "remotes",
    "renv",
    "usethis",

    # Data manipulation & Analysis
    "janitor",
    "rlang",
    "tidyverse",

    # Visualization & Graphics
    "showtext",
    "sysfonts",

    # File & Data management
    "fs",
    "jsonlite",
    "rjson",

    # Document generation & Reporting
    "knitr",
    "pander",
    "quarto",
    "tinytable",

    # Regression & Statistical Modelling
    "lme4",

    # Spatial & Geographic analyses
    "geojsonsf",
    "terra",
    "sdm",

    # Paleoecological/Scientific packages
    "Bchron",
    "neotoma2",

    # Diversity
    "iNEXT",
    "vegan",
    "picante",

    # Phylogenetic analyses,
    "ape",
    "geiger",
    "phytools",

    # Machine learning
    "torch",
    "cito",

    # Datasets
    "palmerpenguins",

    # Development & Programming
    "languageserver",

    # UI & Interactive elements
    "countdown",
    "qrcode"
  )
```

### Test if all packages are ready ✅

Let's do a quick test to make sure everything is in order.
Running the following code should produce `"Everything is good to go!"` instead of an error message.

```{r}
#| label: chunk-pkg-test
#| eval: false
missing_packages <-
  installed.packages() |>
  rownames() |>
  setdiff(x = ssoqe_package_list)

if (
  length(missing_packages) == 0
) {
  cat("Everything is good to go!")
} else {
  paste(
    "The following packages are not installed: \n\n",
    missing_packages |>
      paste(collapse = ", "),
    "\n\nPlease install them before the workshop."
  ) |>
    warning()
}
```

Now, let's install all the missing packages:

```{r}
#| label: chunk-pkg-install
#| eval: false
# Install packages one by one with error handling
for (pkg in missing_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg)
  }
}
```


## Getting the workshop materials ‼️

The internet connection might be an issue at the venue. Therefore we [STRONGLY]{.text-bold} recommend to download ALL workshop materials needed for SSoQE 2025. 

## SSoQE 2025

::: {.callout-important}

The list of required repositories can be found [here](https://github.com/orgs/SSoQE/discussions/16).

:::

## How to download materials?

Here is a step-by-step [guide to download the materials](https://scribehow.com/viewer/Download_a_Reproducibility_in_Science_lecture_from_GitHib__nrjukbSHTpqNcEfliqzTRw). 

Simple instructions:

1. Go to the [SSoQE GitHub repositories](https://github.com/orgs/SSoQE/repositories).
2. click the `Code` button on the right side of the repository page
3. Select `Download ZIP`
4. Unzip the downloaded file to a folder on your computer

### Private repositories

Some materials might not be publicly available and require repository access.

1. Go to [this Discussion thread](https://github.com/orgs/SSoQE/discussions/9) and comment with your GitHub username
2. Wait for organizers to grant you access to the repository
3. Check your email for the invitation to SSoQE
4. You'll then be able to access private repositories and download materials

## Final Checklist Before SSoQE 2025 🎯

Before arriving at the workshop, please verify:

- [ ] R version 4.5+ is installed and working
- [ ] RStudio latest version is installed
- [ ] Rtools installed (Windows users)
- [ ] Git is installed and configured with your name/email
- [ ] GitHub account created
- [ ] All essential packages install without errors
- [ ] Downloaded all required workshop materials

## Need Help? 🆘

If you encounter any issues during setup:

1. Check the [SSoQE Discussion forum](https://github.com/orgs/SSoQE/discussions/categories/questions)
2. Ask questions in the discussion thread
3. We'll help troubleshoot during the workshop introduction

**See you at SSoQE 2025!** 🎉